The Desert of the Reals: Floating-Point Arithmetic on Deterministic Systems

### 2024-04-25 mss.tex
Final cut.

\sloppy
Since the operations and conventions of $\mathbb{R}$ have been found to be so useful, it is desirable to extend the semantics to computer programming.  However, digital computers, by virtue of their binary representation, effectively use natural numbers $\mathbb{N}$ to represent numbers (to the limit of memory rather than positive infinity $+\infty$).  Several schemes permit a computer integer to be interpreted as if it were a number with a fractional part, including a scaling factor, fixed-point representation,\footnote{Hand-in-hand with the development of linear algebra, machines such as {\sc eniac} and {\sc maniac} were employed in the 1940s for solving thermonuclear reaction calculations and neutron diffusion equations.  Under the direction of John von Neumann, it appears that some calculations did experimentally involve a floating-point scheme, although this was later rejected definitively in favor of fixed-point arithmetic.  See \citet{Kahan1997a}, p.~3, on floating-point representations.}

The basic concept of floating point arithmetic is that it permits the representation of a discrete subset of $\mathbb{R}$ by composing a significand, a base, and an exponent.  The significand is the set of significant digits, possibly including the sign; the base is the understood number base (typically 2); and the exponent is the power to which that base is put before multiplying by the significand to yield the result.\footnote{Compare so-called “engineering” notation, such as \texttt{1e5} for 10,000, which compactly represents the significand 1 and the exponent 5 with an understood base of 10 indicated by \texttt{e} and a sign, implicit for positive.}  The most ubiquitous floating-point format in use today is defined by the {\sc ieee} 754 standard, but certain hardware platforms such as {\sc gpu}s utilize alternative floating-point arithmetic representations.\footnote{We cite \texttt{bfloat16} \citep{Wang2019} and TensorFloat-32 \citep{Kharva2020}, among others.}

@@

$$
(-1)^{\texttt{S}} \times 2^{\texttt{E}-127} \times 1.\texttt{F}
$$
\noindent
{\sc ieee} 754 specifies operations between numbers, including of different magnitudes.  The standard dictates behavior and provides outlines for arithmetic, but leaves algorithmic details to the implementation.  Numbers are normalized by adjusting the exponent of the smaller operand and aligning the significands, then the operations are carried out.  In practice, extended precision values are used in the intermediate steps of many algorithms, leading to greater accuracy than would otherwise be expected.\footnote{“Each of the computational operations that return a numeric result specified by this standard shall be performed as if it first produced an intermediate result correct to infinite precision and with unbounded range, and then rounded that intermediate result, if necessary, to fit in the destination’s format.”  \citep{IEEE754-2008}  Note that, per \citeauthor{Risse2016}, “there is no indication whether or not a computation with {\sc ieee} 754 is exact even if all arguments are.”  The {\sc iso}/{\sc iec}9899 C standard confesses its own fallibility:  “The floating-point model is intended to clarify the description of each floating-point characteristic and does not require the floating-point arithmetic of the implementation to be identical” \citetext{\citet{ISO9899}, fn.~21}.}
% https://www.iso.org/obp/ui/#iso:std:iso-iec:9899:ed-4:v1:en

Since the {\sc ieee} 754 floating-point format packs values of different kind together bitwise, conventional integer operations such as left shift $<<$ and addition $+$ do not trivially apply.\footnote{For instance, left-shifting \texttt{1.0} does not result in \texttt{2.0} but \texttt{1.7014118e38}; we leave the mathematics of why as an exercise to the reader.}

@@

\begin{enumerate}
  \item  The distance between two adjacent values changes based on the magnitude of the exponent and the distance from zero.  (The significand resolution stays the same but the exponent changes.)
  \item  There is a relative approximation error for a given bitwidth in {\sc ieee} 754, called the \emph{machine epsilon}.\footnote{Note that this is different from the smallest representable value for a given bit width; e.g., for 32-bit single-precision \texttt{float} the smallest representable value is $\texttt{0000.0000.0000.0000.0000.0000.0000.0001} = 1\times 10^{-45}$.}
  \item  Operations between numbers of different magnitudes are particularly affected by their relative numerical horizon.
\end{enumerate}

@@

Whether something is a gap or optional is something of a philosophical question for us, but some parts are underspecified in a way that makes portability impossible.  E.g., mixed-precision operations can yield unpredictable results depending on the compiler and hardware.  This is a function of rounding modes, precision loss, precision of intermediate results, and the presence or absence of dedicated hardware support for certain precision combinations.

@@

\section{Urbit's Implementation of \textsc{\textbf{ieee}} 754}

@@

Non-real arithmetic is less significant for many of the core operations of Urbit as a personal server platform.  However, gaming, machine learning, graphics, and other applications rely on floating-point calculations—preferably as fast as possible.  In fact, not only applications-oriented processes rely on determinism:  guarantees in cryptography and contractual correctness for web3; verification and validation; accounting and legal compliance; and code correctness analysis all require reproducible determinism.\footnote{Support for {\sc ieee} 754 is similar for support for the Markdown markup language.  Many platforms support a subset of Markdown coupled with platform-specific extensions.  (See also SQL.)  Internal references, HTML, inline \LaTeX~math mode, code block language specification, and other features see varying levels of support with GitHub, Pandoc, Obsidian, and other editors and converters.}

@@

Urbit makes much of avoiding the “ball of mud” “standard software architecture” \citep{Foote1999}.  In this design anti-pattern, a lack of guarantees and predictable behavior leads \emph{inevitably} to haphazard and illegible software bloat \citep{Foote1999}.  We can thus understand why Urbit as a platform considers even deviations in the last bit of a significand to be threads fraying the edge of sanity \citep{Monk2020a}:

@@

For most purposes in the broader software world, tightly reproducible precision has not been a high priority.  Precision having already been sacrificed, the gist of the calculation is more important than the fourth decimal place (e.g. in realtime 3D graphics).  This leads to the phrase “implements the {\sc ieee} 754 standard” being interpreted erroneously to imply full reproducibility \citep{Figueroa2000}.

@@

The situation grows more ambiguous across architectures.  \citetext{\citet{Jones2008}, p.~346,} presents the pathological case of a compliant platform that may use extended precision bits in the calculation of \texttt{a + b}:

@@

K\&R C permitted the compiler to re-order floating-point expressions by associativity, which could run afoul of our limitations.  {\sc ansi} C (C89), recognizing the issue introduced by this innocuous change, forbade such re-ordering \citep{MacDonald1991}.  Compiler optimizations (e.g. {\sc gcc}'s \texttt{-O3}) can bypass this restriction, once again breaking determinism;\footnote{This can be mitigated in turn by the use of the \texttt{volatile} designation, but this is sufficient to illustrate the problem.} for instance, floating-point operations can be pipelined, leading to out-of-order execution.

@@

The challenge of the lack of determinacy for certain critical applications has been acknowledged before, such as by James Demmel and the \texttt{ReproBLAS} team \citep{Demmel2017, ReproBLAS} and by \citeauthor{Dawson2013}.  Dawson makes much of the effect of rounding modes and the option to disable subnormals, both of which would have major effects on computational reproducibility.  The situation is worse for transcendental functions, because there is necessarily truncation and/or rounding error \citep{Dawson2013}.  Even conversion between bases for output and input is not necessarily reproducible, as \citeauthor{Dawson2013} continues:  “Doing perfect conversions efficiently was an unsolved problem when the original {\sc ieee} standard came out, and while it has since been solved, this doesn’t mean that everybody does correctly rounded printing.”  \citet{Koenig2018} expend much effort on the problem of computing a dot product exactly in hardware, given the contingencies of multicore processors.\footnote{The problem of exactness is not exactly the same as reproducibility, but it is related.  Exactness means that the result is the same as if the calculation were carried out to infinite precision and then rounded to the appropriate number of bits.  Reproducibility means that the result is the same across different platforms.  Reproducibitiy requires identical outcomes from the same inputs, while exactness requires the correct answer regardless of algorithmic path.}

@@

Can the C-defined floating-point environment (as supplied by \texttt{fenv.h}) answer to this need?  This affords the ability to specify not only rounding modes and access floating-point exception status flags, but it is not clear whether this environmental control portably spans the entire output of floating-point computations.\footnote{“The floating-point environment access and modification is only meaningful when \texttt{\#pragma STDC FENV\_ACCESS} is set to \texttt{ON}.  \ldots{} In practice, few current compilers, such as HP aCC, Oracle Studio, and IBM XL, support the \texttt{\#pragma} explicitly, but most compilers allow meaningful access to the floating-point environment anyway.” \citep{CPPRef}}

@@

We suggest that deterministically correct stack control in the sense we have described here is impossible for an arbitrary configuration of the modern hardware stack.\footnote{The possibility of circumscribing the set of permissible {\sc ieee} 754 operations, which may afford a different approach to this problem but seems similarly susceptible of shipwreck, is explored in a subsequent section, \emph{quod vide}.}

@@

A related technique could pack bits of larger floating-point values into smaller ones, but this is functionally a software-defined solution (see, e.g., \citet{LeBrun2018}).

@@

In the absence of a dedicated floating-point unit ({\sc fpu}) and floating-point assembly instructions, floating-point computations are carried out in software.  The type can be decomposed from bits, operated on, then packed back into the single type of appropriate value.  For instance, prior to the widespread advent of 64-bit consumer hardware, applications running on PC architecture that needed \texttt{double} values utilized software emulation using two 32-bit numbers together.

@@

While formally correct, software FP is slower than hardware floating point, and likely prohibitively slow for many large matrix applications such as {\sc llm}s.  (“Correctness is more important than performance.”)  Performance is the dolorous stroke against software-defined floating point.  (On the other hand, some early versions of the Apple–IBM–Motorola PowerPC {\sc risc} architecture did not have dedicated hardware floating-point units ({\sc fpu}s) or floating-point assembler instructions at all, requiring full software implementation.\footnote{“There are several 680x0-based Macintosh computers that do not contain floating-point coprocessors” \citep{Apple1994}; on the other hand, “floating-point calculations are performed even faster under the … emulator than on a real 680x0-based Macintosh computer,” indicating that optimized software acceleration is possible, modulo chipset versions and tuned libraries.  The PowerPC 601, introduced in 1991, had a native {\sc fpu}.})  {\sc gcc} has supported a software floating-point mode using \texttt{fp-bit.c} \citep{GCCWiki}; this was particularly used to accommodate the PowerPC limitations rather than to provide either speed or determinism (cf. \citet{Sidwell2006}).

@@

\subsection{Proscribing {\sc ieee} 754}

What if the Scylla of {\sc ieee} 754 is avoided for some other Charybdis?  We can approach this solution space at two levels:  either by sector or entirely.

@@

As an example of a refactoring of {\sc ieee} 754 operations for determinism, consider the \texttt{ReproBLAS} project (last update \textasciitilde 2016.2.21).   \texttt{ReproBLAS} seeks to produce a set of reproducible deterministic algorithms reflecting the standard operations of BLAS \citep{ReproBLAS}.  It accomplishes this by introducing a binned data type and a set of basic operations carefully built on {\sc ieee} 754 for the objective of completely portable reproducibility.\footnote{“Using our default settings, reproducibly summing $n$ floating point types requires approximately $9n$ floating point operations (arithmetic, comparison, and absolute value). In theory, this count can be reduced to $5n$ using the new `augmented addition' and `maximum magnitude' instructions in the proposed {\sc ieee} Floating Point Standard 754-2018.”  \citep{ReproBLAS}}  This is similar to our proposal for a vetted jetting library and may be worth attention, particularly in association with requirements around \texttt{-O0}.

@@

The unum proposal appears to have settled somewhat after its initial state of relative flux (as Type III unums, cf. \citet{Posit2022}).  Although most implementations have been in software, the project has been specified in Verilog several times and implemented on {\sc fpga}s (VividSparks, \url{https://vividsparks.tech/}, and \citet{Chen2018}).  Intriguingly, some initial work has been carried out towards a fundamental {\sc blas}-like library built on posits \citep{vanDam2019}.

@@

A fixed-point representation differs from a floating-point scheme in that the exponent is fixed by the protocol or metadata and thus only the sign and significand need be included in the bit representation.  (With an offset, even the sign can be elided.)  The advantage of such a scheme is that it affords the benefits of floating-point mathematics at near-integer operation speeds (e.g. left-shift to multiply by two).  One disadvantage is that there is a smallest representable value; this lack of subnormals requires either an underflow handler or the possibility of inadvertent division by zero.  Fixed-point operations could also be used as intermediates in calculations.  (This echoes once again the idea of conversion to an intermediate representation then conversion back out to {\sc ieee} 754.)\footnote{“One solution to implementing floating-point types on processors that support fixed-point types is to
convert the source containing floating-point data operations to make calls to a fixed-point library” \citetext{\citet{Jones2008}, p.~346}.  Note that the sense of our current interest is reversed.}

@@

Several recent efforts on Urbit have encountered difficulties in producing reliably deterministic and sufficiently fast floating-point calculations on a Nock-based system.\footnote{Cf. UrWasm, which as a WebAssembly implementation directly relies on floating-point computations \citep{UrWasm} (see e.g. \citet{KloudKoder2022} for a discussion of the issues involved, and in particular \citeauthor{Afonin2024}, pp. 133–150 in this issue), \href{https://github.com/Quodss/urwasm}{\texttt{Quodss/urwasm}}; \texttt{/lib/math}, implemented purely in Hoon, \href{https://github.com/sigilante/libmath}{\texttt{sigilante/libmath}}; and Lagoon, which includes {\sc ieee} 754 operations with its linear algebra library, \href{https://github.com/urbit/numerics}{\texttt{urbit/numerics}} pending development and integration into Urbit.}  We anticipate that, water finding its own level, each will adopt a suitable deterministic solution for evaluation in Nock.  We do not anticipate these to be the last foundational numerical libraries built on Urbit, but instead among the first.  Thus we have documented the paths we have explored as an annotated map for future travelers in search of a one true representation for continuous mathematics. \tombstone{}

>>>>

\sloppy
Since the operations and conventions of $\mathbb{R}$ have been found to be so useful, it is desirable to extend the semantics to computer programming.  However, digital computers, by virtue of their binary representation, effectively use natural numbers $\mathbb{N}$ to represent numbers (to the limit of memory rather than positive infinity $+\infty$).  Several schemes permit a computer integer to be interpreted as if it were a number with a fractional part, including a scaling factor, fixed-point representation,\footnote{Hand-in-hand with the development of linear algebra, machines such as {\sc eniac} and {\sc maniac} were employed in the 1940s for solving thermonuclear reaction calculations and neutron diffusion equations.  Under the direction of John von Neumann, it appears that some calculations did experimentally involve a floating-point scheme, although this was later rejected definitively in favor of fixed-point arithmetic.  See \citet{Kahan1997a}, p.~3, on floating-point representations.} and pairs as rational numbers.

The basic concept of floating point arithmetic is that it permits the representation of a discrete subset of $\mathbb{R}$ by composing a significand, a base, and an exponent.  The significand is the set of significant digits, possibly including the sign; the base is the understood number base (typically 2); and the exponent is the power to which that base is put before multiplying by the significand to yield the result.\footnote{Compare so-called “engineering” notation, such as \texttt{1e5} for 10,000, which compactly represents the significand 1 and the exponent 5 with an understood base of 10 indicated by \texttt{e} and a sign, implicit for positive.}  The most ubiquitous floating-point format today is defined by the {\sc ieee} 754 standard, but certain hardware platforms such as {\sc gpu}s utilize alternative floating-point arithmetic representations.\footnote{We cite \texttt{bfloat16} \citep{Wang2019} and TensorFloat-32 \citep{Kharva2020}, among others.}

@@

$$
(-1)^{\texttt{S}} \times 2^{\texttt{E}-127} \times 1.\texttt{F}
$$
\noindent
{\sc ieee} 754 specifies operations between numbers, including of different magnitudes.  The standard dictates behavior and provides outlines for arithmetic, but leaves algorithmic details to the implementation.  Numbers are normalized by adjusting the exponent of the smaller operand and aligning the significands, then the operations are carried out.  In practice, extended precision values are used in the intermediate steps of many algorithms, leading to greater accuracy than would otherwise be expected.\footnote{“Each of the computational operations that return a numeric result specified by this standard shall be performed as if it first produced an intermediate result correct to infinite precision and with unbounded range, and then rounded that intermediate result, if necessary, to fit in the destination’s format.”  \citep{IEEE754-2008}  Note that, per \citeauthor{Risse2016}, “there is no indication whether or not a computation with {\sc ieee} 754 is exact even if all arguments are.”  The {\sc iso}/{\sc iec} 9899 C standard confesses its own fallibility:  “The floating-point model is intended to clarify the description of each floating-point characteristic and does not require the floating-point arithmetic of the implementation to be identical” \citetext{\citet{ISO9899}, fn.~21}.}
% https://www.iso.org/obp/ui/#iso:std:iso-iec:9899:ed-4:v1:en

Since the {\sc ieee} 754 floating-point format packs values of different kind together bitwise, conventional integer operations such as left shift $<<$ and addition $+$ do not trivially apply.\footnote{For instance, left-shifting a floating-point value does not double it; we leave the mathematics of why as an exercise to the reader.}

@@

\begin{enumerate}
  \item  The distance between two adjacent values changes based on the magnitude of the exponent and the distance from zero.  (The significand resolution stays the same but the exponent changes.)
  \item  There is a relative approximation error for a given bitwidth in {\sc ieee} 754, called the \emph{machine epsilon}.\footnote{Note that this is different from the smallest representable value for a given bit width; e.\,g., for 32-bit single-precision \texttt{float} the smallest representable value is $\texttt{0000.0000.0000.0000.0000.0000.0000.0001} = 1\times 10^{-45}$.}
  \item  Operations between numbers of different magnitudes are particularly affected by their relative numerical horizon.
\end{enumerate}

@@

Whether something is a gap or optional is something of a philosophical question for us, but some parts are underspecified in a way that makes portability impossible.  E.\,g., mixed-precision operations can yield unpredictable results depending on the compiler and hardware.  This is a function of rounding modes, precision loss, precision of intermediate results, and the presence or absence of dedicated hardware support for certain precision combinations.

@@

\section[Urbit's Implementation of \textsc{\textbf{ieee}} 754]{Urbit's Implementation of IEEE 754}

@@

Non-real arithmetic is less significant for many of the core operations of Urbit as a personal server platform.  However, gaming, machine learning, graphics, and other applications rely on floating-point calculations—preferably as fast as possible.  In fact, not only applications-oriented processes rely on determinism:  guarantees in cryptography and contractual correctness for web3; verification and validation; accounting and legal compliance; and code correctness analysis all require reproducible determinism.\footnote{Support for {\sc ieee} 754 is similar for support for the Markdown markup language.  Many platforms support a subset of Markdown coupled with platform-specific extensions.  (See also {\sc sql}.)  Internal references, {\sc html}, inline \LaTeX~math mode, code block language specification, and other features see varying levels of support with GitHub, Pandoc, Obsidian, and other editors and converters.}

@@

\noindent
Urbit makes much of avoiding the “ball of mud” “standard software architecture” \citep{Foote1999}.  In this design anti-pattern, a lack of guarantees and predictable behavior leads \emph{inevitably} to haphazard and illegible software bloat.  We can thus understand why Urbit as a platform considers even deviations in the last bit of a significand to be threads fraying the edge of sanity \citep{Monk2020a}:

@@

For most purposes in the broader software world, tightly reproducible precision has not been a high priority.  Precision having already been sacrificed, the gist of the calculation is more important than the fourth decimal place (e.\,g.\ in realtime 3D graphics).  This leads to the phrase “implements the {\sc ieee} 754 standard” being interpreted erroneously to imply full reproducibility \citep{Figueroa2000}.

@@

The situation grows more ambiguous across architectures.  \citet{Jones2008}, p.~346, presents the pathological case of a compliant platform that may use extended precision bits in the calculation of \texttt{a + b}:

\newpage

@@

K\&R C permitted the compiler to re-order floating-point expressions by associativity, which could run afoul of our limitations.  {\sc ansi} C (C89), recognizing the issue introduced by this innocuous change, forbade such re-ordering \citep{MacDonald1991}.  Compiler optimizations (e.\,g.\ {\sc gcc}'s \texttt{-O3}) can bypass this restriction, once again breaking determinism;\footnote{This can be mitigated in turn by the use of the \texttt{volatile} designation, but this is sufficient to illustrate the problem.} for instance, floating-point operations can be pipelined, leading to out-of-order execution.

@@

The challenge of the lack of determinacy for certain critical applications has been acknowledged before, such as by James Demmel and the \texttt{ReproBLAS} team (\citet{Demmel2017}, \citet{ReproBLAS}) and by \citeauthor{Dawson2013}.  Dawson makes much of the effect of rounding modes and the option to disable subnormals, both of which would have major effects on computational reproducibility.  The situation is worse for transcendental functions, because there is necessarily truncation and/or rounding error \citep{Dawson2013}.  Even conversion between bases for output and input is not necessarily reproducible, as \citeauthor{Dawson2013} continues:  “Doing perfect conversions efficiently was an unsolved problem when the original {\sc ieee} standard came out, and while it has since been solved, this doesn’t mean that everybody does correctly rounded printing.”  \citet{Koenig2018} expend much effort on the problem of computing a dot product exactly in hardware, given the contingencies of multicore processors.\footnote{The problem of exactness is not exactly the same as reproducibility, but it is related.  Exactness means that the result is the same as if the calculation were carried out to infinite precision and then rounded to the appropriate number of bits.  Reproducibility means that the result is the same across different platforms.  Reproducibility requires identical outcomes from the same inputs, while exactness requires the correct answer regardless of algorithmic path.}

@@

Can the C-defined floating-point environment (as supplied by \texttt{fenv.h}) answer to this need?  This affords the ability to specify not only rounding modes and access floating-point exception status flags, but it is not clear whether this environmental control portably spans the entire output of floating-point computations.\footnote{“The floating-point environment access and modification is only meaningful when \texttt{\#pragma STDC FENV\_ACCESS} is set to \texttt{ON}.  \ldots{} In practice, few current compilers, such as HP aCC, Oracle Studio, and {\sc ibm} XL, support the \texttt{\#pragma} explicitly, but most compilers allow meaningful access to the floating-point environment anyway.” \citep{CPPRef}}

@@

We suggest that deterministically correct stack control in the sense we have described here is impossible for an arbitrary configuration of the modern hardware stack.\footnote{The possibility of circumscribing the set of permissible {\sc ieee} 754 operations, which may afford a different approach to this problem but seems similarly susceptible of shipwreck, is explored in a subsequent section, \emph{q.\,v.}}

@@

A related technique could pack bits of larger floating-point values into smaller ones, but this is functionally a software-defined solution (see, e.\,g., \citet{LeBrun2018}).

@@

In the absence of a dedicated floating-point unit ({\sc fpu}) and floating-point assembly instructions, floating-point computations are carried out in software.  The type can be decomposed from bits, operated on, then packed back into the single type of appropriate value.  For instance, prior to the widespread advent of 64-bit consumer hardware, applications that needed \texttt{double} values on 32-bit PC architecture utilized software emulation using two 32-bit numbers together.

@@

While formally correct, software FP is slower than hardware floating point, and likely prohibitively slow for many large matrix applications such as {\sc llm}s.  (“Correctness is more important than performance.”)  Performance is the dolorous stroke against software-defined floating point.  (On the other hand, some early versions of the Apple–{\sc ibm}–Motorola PowerPC {\sc risc} architecture did not have dedicated hardware floating-point units ({\sc fpu}s) or floating-point assembler instructions at all, requiring full software implementation.\footnote{“There are several 680x0-based Macintosh computers that do not contain floating-point coprocessors” \citep{Apple1994}; on the other hand, “floating-point calculations are performed even faster under the … emulator than on a real 680x0-based Macintosh computer,” indicating that optimized software acceleration is possible, modulo chipset versions and tuned libraries.  The PowerPC 601, introduced in 1991, had a native {\sc fpu}.})  {\sc gcc} has supported a software floating-point mode using \texttt{fp-bit.c} \citep{GCCWiki}; this was particularly used to accommodate the PowerPC limitations rather than to provide either speed or determinism (cf. \citet{Sidwell2006}).

@@

\subsection[Proscribing {\sc ieee} 754]{Proscribing IEEE 754}

What if the Scylla of {\sc ieee} 754 is avoided for some Charybdis?  We can approach this solution space at two levels:  either by sector or entirely.

@@

As an example of a refactoring of {\sc ieee} 754 operations for determinism, consider the \texttt{ReproBLAS} project (last update \textasciitilde 2016.2.21).   \texttt{ReproBLAS} seeks to produce a set of reproducible deterministic algorithms reflecting the standard operations of {\sc blas} \citep{ReproBLAS}.  It accomplishes this by introducing a binned data type and a set of basic operations carefully built on {\sc ieee} 754 for the objective of completely portable reproducibility.\footnote{“Using our default settings, reproducibly summing $n$ floating point types requires approximately $9n$ floating point operations (arithmetic, comparison, and absolute value). In theory, this count can be reduced to $5n$ using the new `augmented addition' and `maximum magnitude' instructions in the proposed {\sc ieee} Floating Point Standard 754-2018.”  \citep{ReproBLAS}}  This is similar to our proposal for a vetted jetting library and may be worth attention, particularly in association with requirements around \texttt{-O0}.

@@

The unum proposal appears to have settled somewhat after its initial state of relative flux (as Type {\sc iii} unums, cf. \citet{Posit2022}).  Although most implementations have been in software, the project has been specified in Verilog several times and implemented on {\sc fpga}s (\citet{Chen2018}, and VividSparks, \url{https://vividsparks.tech/}).  Intriguingly, some initial work has been carried out towards a fundamental {\sc blas}-like library built on posits \citep{vanDam2019}.

@@

A fixed-point representation differs from a floating-point scheme in that the exponent is fixed by the protocol or metadata and thus only the sign and significand need be included in the bit representation.  (With an offset, even the sign can be elided.)  The advantage of such a scheme is that it affords the benefits of floating-point mathematics at near-integer operation speeds (e.\,g.\ left-shift to multiply by two).  One disadvantage is that there is a smallest representable value; this lack of subnormals requires either an underflow handler or the possibility of inadvertent division by zero.  Fixed-point operations could also be used as intermediates in calculations.  (This echoes once again the idea of conversion to an intermediate representation then conversion back out to {\sc ieee} 754.)\footnote{“One solution to implementing floating-point types on processors that support fixed-point types is to
convert the source containing floating-point data operations to make calls to a fixed-point library” \citetext{\citet{Jones2008}, p.~346}.  Note that the sense of our current interest is reversed.}

@@

Several recent efforts on Urbit have encountered difficulties in producing reliably deterministic and sufficiently fast floating-point calculations on a Nock-based system.\footnote{Cf. UrWasm, which as a WebAssembly implementation directly relies on floating-point computations \citep{UrWasm} (see e.\,g.\ \citet{KloudKoder2022} for a discussion of the issues involved, and in particular \citeauthor{Afonin2024}, pp. 133–150 in this issue), \href{https://github.com/Quodss/urwasm}{\texttt{Quodss/urwasm}}; \texttt{/lib/math}, implemented purely in Hoon, \href{https://github.com/sigilante/libmath}{\texttt{sigilante/libmath}}; and Lagoon, which includes {\sc ieee} 754 operations with its linear algebra library, \href{https://github.com/urbit/numerics}{\texttt{urbit/numerics}} pending development and integration into Urbit.}  We anticipate that, water finding its own level, each will adopt a suitable deterministic solution for evaluation in Nock.  We do not anticipate these to be the last foundational numerical libraries built on Urbit, but instead among the first.  Thus we have documented the paths we have explored as an annotated map for future travelers in search of a one true representation for continuous mathematics. \tombstone{}
